### 2023-01-26 Strydom_et_al_2023_graph_embedding_draft.tex
Deploying to gh-pages from @ PoisotLab/ms_metaweb_perspectives@a5efe396f635fafb5ba5f2ef1440b056bf76d3a7 🚀

Being able to infer \emph{potential} interactions could serve as a
significant breakthrough in our ability to start thinking about species
interaction networks over large spatial scales (Hortal et al., 2015).
Understanding species interactions holds enormous potential to not only
understand and more rapidly learn about species interactions and
metawebs, but also how changes in management of a single species may
impact non-target species. In a recent overview of the field of
ecological network prediction, Strydom, Catchen, et al. (2021)
identified two challenges of interest to the prediction of interactions
at large scales. First, there is a relative scarcity of relevant data in
most places globally -- paradoxically, this restricts our ability to
infer interactions to locations where inference is perhaps the least
required (and leaves us unable to make inference in data scarce
regions); second, accurate predictors are important for accurate
predictions, and the lack of methods that can leverage a small amount of
\emph{accurate} data is a serious impediment to our predictive ability.
In most places, our most reliable biodiversity knowledge is that of a
species pool (\emph{i.e.} a set of potentially interacting species in a
given area): through the analysis of databases like the Global
Biodiversity Information Facility (GBIF) or the International Union for
the Conservation of Nature (IUCN), it is possible to construct a list of
species for a region of interest; however inferring the potential
interactions between these species still remains a challenge.

@@

@@

@@

\hypertarget{the-metaweb-embeds-both-ecological-hypotheses-and-practices}{%
\section{The metaweb embeds both ecological hypotheses and
practices}\label{the-metaweb-embeds-both-ecological-hypotheses-and-practices}}

The goal of metaweb inference is to provide information about the
interactions between species at a large spatial scale. But as Herbert
(1965) rightfully pointed out, ``{[}y{]}ou can't draw neat lines around
planet-wide problems''; any inference of a metaweb at large scales must
contend with several novel, and interwoven, families of problems. In
this section, we list some of the most pressing research priorities
(\emph{i.e.} problems that can be addressed with subsequent data
analysis or simulations), as well as issues related to the application
of these methods at the science-policy interface.

@@

If the initial metaweb is too narrow in scope, notably from a taxonomic
point of view, the chances of finding another area with enough related
species (through phylogenetic relatedness or similarity of functional
traits) to make a reliable inference decreases; this would likely be
indicated by large confidence intervals during estimation of the values
in the low-rank space, meaning that the representation of the original
graph is difficult to transfer to the new problem. Alternatively, if the
initial metaweb is too large (taxonomically), then the resulting
embeddings would need to represent interactions between taxonomic groups
that are not present in the new location. This would lead to a much
higher variance in the starting dataset, and to under-dispersion in the
target dataset, resulting in the potential under or over estimation of
the strength of new predicted interactions. The lack of well documented
metawebs is currently preventing the development of more concrete
guidelines. The question of phylogenetic relatedness and distribution is
notably relevant if the metaweb is assembled in an area with mostly
endemic species (\emph{e.g.} a system that has undergone recent
radiation or that has remained in isolation for a long period of time
might not have an analogous system with which to draw knowledge from),
and as with every predictive algorithm, there is room for the
application of our best ecological judgement. Because this problem
relates to distribution of species in the geographic or phylogenetic
space, it can certainly be approached through assessing the performance
of embedding transfer in simulated starting/target species pools.

@@

The area used to infer the new metaweb in determines the species pool
that must be used to perform this embedding. Metawebs can be constructed
by assigning interactions in a list of species within geographic
boundaries. The upside of this approach is that information at the
country level is likely to be required for biodiversity assessments, as
countries set conservation goals at the national level (Buxton et al.,
2021), and as quantitative instruments are designed to work at these
scales (Turak et al., 2017); specific strategies are often enacted at
smaller scales, nested within a specific country (Ray et al., 2021). But
there is no guarantee that these boundaries are meaningful. In fact, we
do not have a satisfying answer to the question of ``where does an
ecological network stop?''. Recent results by Martins et al. (2022)
suggest that networks are shaped within eco-regions, with abrupt
structural transitions from an eco-region to the next. Should this trend
hold generally, this would provide an ecologically-relevant scale at
which metawebs can be downscaled and predicted. Other solutions leverage
network-area relationships to identify areas in which networks are
structurally similar (see \emph{e.g.} Fortin et al., 2021; Galiana et
al., 2022, 2018), which requires ample pre-existing information about
the network. This suggests that inferred metawebs should be further
downscaled to allow for \emph{a posteriori} analyses; Llewelyn et al.
(2022) provide compelling evidence for the fact that when a place is
rich in endemic species, even at smaller spatial scales, the transfer of
information about interactions becomes more challenging, re-emphasizing
the need to identify guidelines for when and where network predictions
can be mapped.

@@

Operating under the context of national divisions, in large parts of the
world, reflects nothing more than the legacy of settler colonialism,
which drives a disparity in available ecological data. Applying any
embedding to biased data does not debias them, but instead embeds these
very same biases, propagating them to the machine learning models using
embeddings to make predictions. Indeed, the use of ecological data
itself is not an apolitical act (Nost \& Goldstein, 2021), as data
infrastructures tend to be designed to answer questions within national
boundaries (therefore placing contingencies on what is available to be
embedded). Furthermore, their use often draws upon, and reinforces,
territorial statecraft (see \emph{e.g.} Barrett, 2005). As per Machen \&
Nost (2021), these biases are particularly important to consider when
the output of ``algorithmic thinking'' (\emph{i.e.} relying on machine
learning to generate knowledge or as a substitute to human
decision-making) can be re-purposed for governance (\emph{e.g.} enacting
conservation decisions on the basis of model prediction). As information
on species interaction networks structure is increasingly leveraged as a
tool to guide conservation actions (see \emph{e.g.} Eero et al., 2021;
Naman et al., 2022; Stier et al., 2017), the need to appraise and
correct biases that are unwittingly propagated to algorithms when
embedded from the original data is paramount. These considerations are
even more urgent in the specific context of biodiversity data, as
long-term colonial legacies still shape taxonomic composition to this
day (Lenzner et al., 2022; Raja, 2022), and where much shorter-term
changes in taxonomic and genetic richness of wildlife emerged through
environmental racism (Schmidt \& Garroway, 2022).

\hypertarget{conclusion-metaweb-prediction-in-context}{%
\section{Conclusion: metaweb prediction in
context}\label{conclusion-metaweb-prediction-in-context}}

Predictive approaches, regardless of the scale at which they are
deployed and the intent of their deployment, originate in the framework
that contributed to the ongoing biodiversity crisis (Adam, 2014) and
reinforced environmental injustice (Choudry, 2013; Domínguez \& Luoma,
2020). The risk of embedding this legacy in our models is great,
especially when the impact of this legacy on species pools is being
increasingly documented. This problem can be addressed by re-framing the
way we interact with models, especially in the context where these
models are intended to support conservation actions. Particularly on
territories that were traditionally stewarded by Indigenous people,
these approaches should be replaced (or at least guided and framed) by
Indigenous principles of land management (Eichhorn et al., 2019; No'kmaq
et al., 2021), as part of an ``algorithm-in-the-loop'' approach.
Human-algorithm interactions are notoriously difficult and can yield
adverse effect (Green \& Chen, 2019; Stevenson \& Doleac, 2021),
suggesting the need to systematically study them for the specific
purpose of biodiversity governance, as well as to improve the
algorithmic literacy of decision makers. As we see artificial
intelligence/machine learning being increasingly mobilized to generate
knowledge that is lacking for conservation decisions (\emph{e.g.} Lamba
et al., 2019; Mosebo Fernandes et al., 2020) and drive policy decisions
(Weiskopf et al., 2022), our discussion of these tools need to go beyond
the technical and statistical, and into the governance consequences they
can have.

>>>>

The ability to infer \emph{potential} interactions could serve as a
significant breakthrough in our ability to conceptualize species
interaction networks over large spatial scales (Hortal et al., 2015).
Robust inferences would not only boost our understanding of the
structure of species interaction networks, but also increase the amount
of information that can be used for biodiversity management. In a recent
overview of the field of ecological network prediction, Strydom,
Catchen, et al. (2021) identified two challenges of interest to the
prediction of interactions at large scales. First, there is a relative
scarcity of relevant data in most places globally -- paradoxically, this
restricts our ability to infer interactions to locations where inference
is perhaps the least required (and leaves us unable to make inference in
data scarce regions); second, accurate predictors are important for
accurate predictions, and the lack of methods that can leverage a small
amount of \emph{accurate} data is a serious impediment to our predictive
ability. In most places, our most reliable biodiversity knowledge is
that of a species pool (\emph{i.e.} a set of potentially interacting
species in a given area): through the analysis of databases like the
Global Biodiversity Information Facility (GBIF) or the International
Union for the Conservation of Nature (IUCN), it is possible to construct
a list of species for a region of interest; however inferring the
potential interactions between these species still remains a challenge.

@@

\hypertarget{ecological-networks-are-good-candidates-for-embedding}{%
\subsection{Ecological networks are good candidates for
embedding}\label{ecological-networks-are-good-candidates-for-embedding}}

@@

\hypertarget{graph-embeddings-produce-novel-latent-variables-to-predict-interactions}{%
\subsection{Graph embeddings produce novel latent variables to predict
interactions}\label{graph-embeddings-produce-novel-latent-variables-to-predict-interactions}}

\hypertarget{graph-embedding-has-been-under-used-in-the-prediction-of-species-interactions}{%
\subsection{Graph embedding has been under-used in the prediction of
species
interactions}\label{graph-embedding-has-been-under-used-in-the-prediction-of-species-interactions}}

@@

\hypertarget{the-metaweb-merges-ecological-hypotheses-and-practices}{%
\section{The metaweb merges ecological hypotheses and
practices}\label{the-metaweb-merges-ecological-hypotheses-and-practices}}

Metaweb inference seeks to provide information about the interactions
between species at a large spatial scale, typically a scale large enough
to be considered of biogeographic relevance (indeed, many of the
examples covered in the introduction span areas larger than a country,
some of them global). But as Herbert (1965) rightfully pointed out,
``{[}y{]}ou can't draw neat lines around planet-wide problems''; any
inference of a metaweb must therefore contend with several novel,
interwoven, families of problems. In this section, we outline three that
we think are particularly important, and can discuss how they may
addressed with subsequent data analysis or simulations, and how they
emerge in the specific context of using embeddings; some of these issues
are related to the application of these methods at the science-policy
interface.

@@

If the initial metaweb is too narrow in scope, notably from a taxonomic
point of view, the chances of finding another area with enough related
species (through phylogenetic relatedness or similarity of functional
traits) to make a reliable inference decreases. This is because transfer
requires similarity (fig.~\ref{fig:embedding}). A diagnostic for the
lack of similar species would likely be large confidence intervals
during estimation of the values in the low-rank space. In other words,
the representation of the original graph is difficult to transfer to the
new problem. Alternatively, if the initial metaweb is too large
(taxonomically), then the resulting embeddings would need to represent
interactions between taxonomic groups that are not present in the new
location. This would lead to a much higher variance in the starting
dataset, and to under-dispersion in the target dataset, resulting in the
potential under or over estimation of the strength of new predicted
interactions. Llewelyn et al. (2022) provide compelling evidence for
these situations by showing that, even at small spatial scales, the
transfer of information about interactions becomes more challenging when
areas rich with endemic species are considered. The lack of well
documented metawebs is currently preventing the development of more
concrete guidelines. The question of phylogenetic relatedness and
distribution is notably relevant if the metaweb is assembled in an area
with mostly endemic species (\emph{e.g.} a system that has undergone
recent radiation or that has remained in isolation for a long period of
time might not have an analogous system with which to draw knowledge
from), and as with every predictive algorithm, there is room for the
application of our best ecological judgement. Because this problem
relates to distribution of species in the geographic or phylogenetic
space, it can certainly be approached through assessing the performance
of embedding transfer in simulated starting/target species pools.

@@

The area for which we seek to predict the metaweb should determine the
species pool on which the embedding is performed. Metawebs can be
constructed by assigning interactions in a list of species within
geographic boundaries. The upside of this approach is that information
relevant for the construction of this dataset is likely to exist, as
countries usually set conservation goals at the national level (Buxton
et al., 2021), and as quantitative instruments are consequently designed
to work at these scales (Turak et al., 2017); specific strategies are
often enacted at smaller scales, nested within a specific country (Ray
et al., 2021). But there is no guarantee that these boundaries are
meaningful. In fact, we do not have a satisfying answer to the question
of ``where does an ecological network stop?'', the answer to which would
dictate the spatial span to embed/predict. Recent results by Martins et
al. (2022) suggest that networks are shaped within eco-regions, with
abrupt structural transitions from an eco-region to the next. Should
this trend hold generally, this would provide an ecologically-relevant
scale at which metawebs can be downscaled and predicted. Other solutions
could leverage network-area relationships to identify areas in which
networks are structurally similar (see \emph{e.g.} Fortin et al., 2021;
Galiana et al., 2022, 2018). Both of these solutions requires ample
pre-existing information about the network in space. Nevertheless, the
inclusion of species for which we have data but that are not in the
right spatial extent\emph{may} improve the performance of approaches
based on embedding and transfer, \emph{if} they increase the similarity
between the target and destination network. This proposal can
specifically be evaluated by adding nodes to the network to embed, and
evaluating the performance of predictive models (see \emph{e.g.}
Llewelyn et al., 2022).

@@

In large parts of the world, geographic boundaries only reflect the
legacy of settler colonialism, which drives global disparity in capacity
to collect and publish ecological data. Applying any embedding to biased
data does not debias them, but rather embeds these biases, propagating
them to the models using embeddings to make predictions. Furthermore,
the use of ecological data itself is not an apolitical act (Nost \&
Goldstein, 2021): data infrastructures tend to be designed to answer
questions within national boundaries (therefore placing contingencies on
what is available to be embedded), their use often drawing upon, and
reinforcing, territorial statecraft (see \emph{e.g.} Barrett, 2005). As
per Machen \& Nost (2021), these biases are particularly important to
consider when knowledge generated algorithmically is used to supplement
or replace human decision-making, especially for governance (\emph{e.g.}
enacting conservation decisions on the basis of model prediction). As
information on networks is increasingly leveraged for conservation
actions (see \emph{e.g.} Eero et al., 2021; Naman et al., 2022; Stier et
al., 2017), the need to appraise and correct biases that are unwittingly
propagated to algorithms when embedded from the original data is
immense. These considerations are even more urgent in the specific
context of biodiversity data. Long-term colonial legacies still shape
taxonomic composition to this day (Lenzner et al., 2022; Raja, 2022),
and much shorter-term changes in taxonomic and genetic richness of
wildlife emerged through environmental racism (Schmidt \& Garroway,
2022). These considerations mean that ``the set of species found at a
place'' is not only a response to ecological processes, and therefore
there may exist forcing on the interactions between these species.

\hypertarget{conclusion-metawebs-predictions-and-people}{%
\section{Conclusion: metawebs, predictions, and
people}\label{conclusion-metawebs-predictions-and-people}}

Predictive approaches in ecology, regardless of the scale at which they
are deployed and the intent of their deployment, originate in the
framework that contributed to the ongoing biodiversity crisis (Adam,
2014) and reinforced environmental injustice (Choudry, 2013; Domínguez
\& Luoma, 2020). The risk of embedding this legacy in our models is
real, especially when the impact of this legacy on species pools is
being increasingly documented. This problem can be addressed by
re-framing the way we interact with models, especially when models are
intended to support conservation actions. Particularly on territories
that were traditionally stewarded by Indigenous people, we must
interrogate how predictive approaches and the biases that underpin them
can be put to task in accompanying Indigenous principles of land
management (Eichhorn et al., 2019; No'kmaq et al., 2021). The discussion
of ``algorithm-in-the-loop'' approaches that is now pervasive in the
machine learning community provides examples of why this is important.
Human-algorithm interactions are notoriously difficult and can yield
adverse effect (Green \& Chen, 2019; Stevenson \& Doleac, 2021),
suggesting the need to systematically study them for the specific
purpose of, here, biodiversity governance. Improving the algorithmic
literacy of decision makers is part of the solution (\emph{e.g.} Lamba
et al., 2019; Mosebo Fernandes et al., 2020), as we can reasonably
expect that model outputs will be increasingly used to drive policy
decisions (Weiskopf et al., 2022). Our discussion of these approaches
need to go beyond the technical and statistical, and into the governance
consequences they can have. To embed data also embeds historical and
contemporary biases that acted on these data, both because they shaped
the ecological processes generating them, and the global processes
leading to their measurement and publication. For a domain as vast as
species interaction networks, these biases exist at multiple scales
along the way, and a challenge for prediction is not only to develop (or
adopt) new quantitative tools, but to assess the behavior of these tools
in the proper context.
